\chapter{Conclusions}
\label{cha:conclusions}

This thesis described the design and implementation of an asynchronous data
access middleware for Pervasive Systems. As shown in previous chapters, this
process began with an analysis aimed at identifying the weaknesses and
strengths of the Classic PerLa Middleware architecture, which was later used to
outline a basic set of goals to be followed during the development of the
software hereby described. From these goals ensued a New Middleware design
(chapter~\ref{cha:middleware_overview}), and a concrete implementation
(chapter~\ref{cha:components}). The most important contributions that the
development of this new data access middleware brought to the PerLa System can
be classified into three categories: modularization of the Plug \& Play device
registration process, asynchronous data flow management, and an improved
\texttt{FPC}. 

The new Plug \& Play device registration process was enhanced with three
distinct measures: first, the internal structure of the \texttt{FPC} component
was split into independent modules; second, the \texttt{FPCFactory} itself was
partitioned into several sub-factory units, one for each \texttt{FPC} module;
third, a Plugin System was designed to allow the addition of new \texttt{FPC}
fragments without requiring any direct modification to the \texttt{FPCFactory}
itself. The advantages and merits of this new modular design were tested and
validated with the implementation of five different modules, three created by
the author of this thesis (\texttt{JSONMapper}, \texttt{URLEncodedMapper} and
\texttt{SimulatorChannel}), and two by other graduate student
(\texttt{HTTPChannel} and \texttt{TinyOSChannel}).

Another crucial aspect of the New PerLa Middleware design is represented by the
asynchronous data flow management paradigm. As explained in
section~\ref{sec:newmiddleware.async}, all components of the new architecture
implement an asynchronous event-driven API that improves both memory usage and
global reaction times of the entire system. An example of the benefits brought
by this new paradigm can be derived by analysing the number of threads
instantiated for each \texttt{FPC}. In the Classic Middleware, each
\texttt{FPC} was composed of four distinct Java threads: one dedicated to
reading incoming messages, one for the \texttt{Unmarshaller}, one for the
\texttt{Marshaller}, and one for the creation of output records. Conversely,
within the New Middleware infrastructure, a single Java thread located in the
\texttt{Channel} module is enough to drive all data handoffs occurring inside
an entire \texttt{FPC}. Although additional threads may be instantiated during
the execution of PerLa \texttt{Script}s, it is worth noting that these are
managed by the \texttt{Script Engine}, and are always shared among all running
\texttt{FPC}s; as a consequence, the New Middleware can dynamically adjust its
resource consumption figures to match the actual workload (see
chapter~\ref{cha:components} for additional details).

The \texttt{FPC} benefits from another improvement brought by the New
Middleware design, namely the PerLa Scripting Language. This new feature
complements the declarative nature of the Device Descriptor, enabling the
definition of advanced mappings between device capabilities and data
\texttt{Attributes} exposed by the \texttt{FPC} component. The imperative
programming paradigm fostered by the PerLa Scripting Language proved to be a
key addition to the Middleware architecture, as it enhanced the flexibility and
versatility of the entire PerLa System; \texttt{Script}s have in fact been used
to define complex device initialization procedures, to aggregate the
information collected from a sensing network, and to reshape the contents of
hierarchical data structures into the one-dimensional record pattern produced
by the \texttt{FPC} component. It is in the author's view that the former
Device Descriptor structure could not be used to adequately model the
aforementioned applications, as its inherently declarative essence embodied a
fixed set of behavioural assumption which were forced on all nodes of the
Pervasive Network.  The PerLa Scripting Language proposes itself as a less
opinionated tool that can be used by node developers to better specify the
functioning mechanisms of their devices.


\section{Future work}

\subsection{Implementation of new plugins}

The New PerLa Middleware is designed to be extended through the addition of new
modules, and it should come as no surprise that one of its intended evolution
paths consists in fact in the development of new Plugins. As described in
chapter~\ref{cha:components}, there are two main types of modules that can be
added to the PerLa Plugin System: \texttt{Channels} and \texttt{Mappers}.

The possibility to add new \texttt{Channel} implementations is a distinguishing
feature of the New Middleware design that should be effectively exploited to
expand the range of supported endpoint devices. At the time of writing the
selection of Plugins shipped with the core Middleware distribution allows PerLa
to connect with HTTP services and TinyOS motes. This initial line-up should be
only considered as a starting point, since a larger assortment of communication
systems is required to manage even the most rudimentary Pervasive System. The
following list contains a choice of protocols and networking technologies for
which a dedicated \texttt{Channel} implementation is advised:

\begin{itemize}

    \item \textbf{TCP/IP:} Widely employed in a vast variety of devices,
    ranging from high-end personal computers to low power devices;

\item \textbf{Bluetooth LE (Low Energy):} One of the leading technologies for
    wireless personal area networks. Currently supported by all major operating
    systems, Bluetooth LE found its way into many devices and appliances, like
    fitness bands, smartphones, healthcare instruments, home entertainment sets
    and localization beacons;

    \item \textbf{IEEE 802.15.4 based protocols:} IEEE 802.15.4 is a physical
    layer widely employed in many personal area network protocols. It is the
    foundation of several networking specifications like Zigbee, Xbee and MiWi;

    \item \textbf{RS232:} Serial port communication. Its implementation should
    be considered in order to connect with legacy devices and other low power
    systems (e.g., Arduino).

\end{itemize}

\texttt{Mapper}s represent another extension point of the PerLa Middleware
infrastructure that can be used to manage additional data formats and
encodings. The only \texttt{Mapper} components available as of December 2014 in
the core Middleware distribution provide support for JSON and URL-encoded data
structures. Analogously to what already stated for the \texttt{Channel}
component, future PerLa developers should consider implementing new
\texttt{Mapper}s for handling the following data formats:

\begin{itemize}

    \item \textbf{C/C++ structs:} Its implementation should be a simple
    backport from the Classic Middleware architecture;

    \item \textbf{XML:} A markup language for document encoding;

    \item \textbf{CSV (Comma-Separated Values):} A simple format for the
    transmission and storage of tabular data.

\end{itemize}

\subsection{Alternative Device Descriptor forms}

As introduced in chapter~\ref{cha:components}, the new Plug \& Play node
registration system is comprised of two separate elements: a
\texttt{DeviceDescriptorParser} front-end, which analyzes the Device Descriptor
to build a format-agnostic Java representation of the descriptor itself, and
the \texttt{FPCFactory}, which consumes this intermediate Java representation
to assemble the final \texttt{FPC}. This new architecture was conceived to
facilitate the future addition of alternative Device Descriptor formats by only
requiring the development of an appropriate \texttt{DeviceDescriptorParser}
module.

There are two main reasons for adding a new Device Descriptor representation:
first, to support a different data format that may be more convenient for some
devices (e.g., a JSON Device Descriptor); second, to create \texttt{FPC}s using
readily available industry standard device description technologies. This
former motivation leads us to one future development of the PerLa
infrastructure, namely the possibility of introducing a
\texttt{DeviceDescriptorParser} for the \textit{SensorML} \cite{sensorml}
sensor description format. Through this effort the PerLa Middleware would be
immediately compatible with all devices for which a SensorML description is
already available.

\subsection{Distributed PerLa}

Future development of the PerLa Middleware should aim at implementing
in-network processing capabilities in order to better exploit the resources
available in a Pervasive System. Such efforts must focus on the definition of a
software distribution infrastructure that can be used to divide a high-level
computation into smaller, independent units of work to be executed on the
individual nodes of the sensing network.
